### 2014-11-12 paper/paths.tex
through path type preservation

\section*{Types}

Values in Patina can be described by the following types:
\[
\begin{array}{lccl}
\textrm{Lifetime} & \lt & & \\
\textrm{Type Variable} & X & & \\
\textrm{Qualifier} & q & \bnfdef & \qimm \bnfalt \qmut \\
\textrm{Type} & \gt & \bnfdef & \tyint \bnfalt \own{\gt} \bnfalt \tyref{\lt}{q}{\gt} \bnfalt 
				\subvar{\gt}{i}{n} \bnfalt \subrec{\gt}{i}{n} \bnfalt 
				\tyfix{X}{\gt} \bnfalt X \\
\end{array}
\]

$\own{\gt}$ is a unique pointer to a $\gt$.
$\tyref{\lt}{q}{\gt}$ is a borrowed reference to a $\gt$
providing mutability guarantee $q$ for lifetime $\lt$.
$\subvar{\gt}{i}{n}$ is a variant type.
$\subrec{\gt}{i}{n}$ is a tuple type.
Finally, $\tyfix{X}{\gt}$ is a recursive type.

@@

>>>>

@@

\section*{Runtime Typing}
\subsection*{Route Typing}
Typing routes is much like typing paths; however, the $\pay{r}$ route requires
special attention. Since this route points to the payload of a variant,
its type depends upon the value of the discriminant of the variant.
We handle this with an additional context that records the discriminant value
of a variant stored at a route.

$$ \gS : \mathrm{Allocation} \to \mathrm{Type}$$
$$ \gY : \mathrm{Allocation} \times \mathrm{Route} \to \mathrm{Integer} $$

\fbox{$\tc{\gS;\gY}{\addr{\ga}{r}}{\gt} $}

\begin{mathpar}
\infer[RT-BASE]{\gS(\ga)=\gt}{\tc{\gS;\gY}{\addr{\ga}{\base}}{\gt}}
\and
\infer[RT-PROJ]
  {\tc{\gS;\gY}{\addr{\ga}{r}}{\subrec{\gt}{i}{n}}}
  {\tc{\gS;\gY}{\addr{\ga}{\proj{r}{i}}}{\gt_i}}
\and
\infer[RT-PAY]
  {\gY(\ga,r)=i \\ \tc{\gS;\gY}{\addr{\ga}{r}}{\subvar{\gt}{i}{n}}}
  {\tc{\gS;\gY}{\addr{\ga}{\pay{r}}}{\gt_i}}
\and
\infer[RT-UNROLL]
  {\tc{\gS;\gY}{\addr{\ga}{r}}{\tyfix{X}{\gt}}}
  {\tc{\gS;\gY}{\addr{\ga}{\unroll{\tyfix{X}{\gt}}{r}}}{\sub{X}{\tyfix{X}{\gt}}{\gt}}}
\end{mathpar}

\subsection*{Layout and Cell Typing}
Typing for layouts and cells is what you would expect.
However, we restrict the type of $\void$ to the possible types for other
kinds of cells, i.e. $\tyint$, unique pointers, and borrowed references.
This means that multi-cell layout types (variants, tuples, and recursive types) imply
a non-$\void$ value.
\newline

\fbox{$\tc{\gS;\gY}{l}{\gt}$}

\begin{mathpar}
\infer[LT-VOIDINT]
{ }
{\tc{\gS;\gY}{\void}{\tyint}}
\and
\infer[LT-VOIDOWN]
{ }
{\tc{\gS;\gY}{\void}{\own{\gt}}}
\and
\infer[LT-VOIDREF]
{ }
{\tc{\gS;\gY}{\void}{\tyref{\lt}{q}{\gt}}}
\and
\infer[LT-INT]
{ }
{\tc{\gS;\gY}{z}{\tyint}}
\and
\infer[LT-OWN]
{\tc{\gS;\gY}{\addr{\ga}{r}}{\gt}}
{\tc{\gS;\gY}{\own{\addr{\ga}{r}}}{\own{\gt}}}
\and
\infer[LT-REF]
{\tc{\gS;\gY}{\addr{\ga}{r}}{\gt}}
{\tc{\gS;\gY}{\refval{q}{\addr{\ga}{r}}}{\tyref{\lt}{q}{\gt}}}
\and
\infer[LT-VOIDVAR]
{ }
{\tc{\gS;\gY}{\varval{\void}{\void}}{\subvar{\gt}{i}{n}}}
\and
\infer[LT-VAR]
{\tc{\gS;\gY}{l}{\gt_i}}
{\tc{\gS;\gY}{\varval{i}{l}}{\subvar{\gt}{i}{n}}}
\and
\infer[LT-REC]
{\forall i.~\tc{\gS;\gY}{l_i}{\gt_i}}
{\tc{\gS;\gY}{\subrec{l}{i}{n}}{\subrec{\gt}{i}{n}}}
\and
\infer[LT-ROLL]
{\tc{\gS;\gY}{l}{\sub{X}{\tyfix{X}{\gt}}{\gt}}}
{\tc{\gS;\gY}{\roll{\tyfix{X}{\gt}}{l}}{\tyfix{X}{\gt}}}
\end{mathpar}

\subsection*{Heap Well-Formedness}
With runtime typing now in hand, we can say what it means for the heap to be well-formed.
We say a heap $H$ is described by a heap type $\gS$ and a discriminant context $\gY$ if
every allocation in $H$ has the type assigned to it by $\gS$
and $\gY$ correctly records the discrimiants of all the variants in $H$.
Formally,

\begin{mathpar}
\mprset{flushleft}
\infer
{ 
\dom{H} = \dom{\gS} \\\\
\forall \ga \in \dom(\gS).~\gS(\ga)~\textrm{closed} \\\\
\forall \ga \in \dom{H}.~\tc{\gS;\gY}{H(\ga)}{\gS(\ga)} \\\\
\forall (\ga,r) \in \dom{\gY}.~\Read{H}{\addr{\ga}{r}}{\varval{\gY(\ga,r)}{l}}
}
{\tc{}{H}{\gS;\gY}}
\end{mathpar}

\section*{Read Safety}
We are now in position to prove that our read operation is safe.
That is, if the heap is well formed and a route is has some type,
then we can read some layout at that route and that layout has that same type.

\begin{lem}[Read Safety]
If $\tc{}{H}{\gS;\gY}$ and $\tc{\gS;\gY}{\addr{\ga}{r}}{\gt}$,
then there is a layout $l$ such that $\Read{H}{\addr{\ga}{r}}{l}$ and $\tc{\gS;\gY}{l}{\gt}$.
\end{lem}

\begin{proof}[Read Safety]
  We will use induction on the derivation of $\tc{\gS;\gY}{\addr{\ga}{r}}{\gt}$.
  \begin{itemize}
    \item[\textsc{RT-BASE}] Then $r = \base$.

      By inversion, $\gS(\ga)=\gt$.

      From $\tc{}{H}{\gS;\gY}$, we know that $\dom{H}=\dom{\gS}$.
      Thus, $\exists l.~H(\ga)=l$.

      Then by \textsc{RD-BASE}, we have $\Read{H}{\addr{\ga}{\base}}{l}$.

      Again from $\tc{}{H}{\gS;\gY}$, we know that $\tc{\gS;\gY}{H(\ga)}{\gS(\ga)}$,
      which is just $\tc{\gS;\gY}{l}{\gt}$ as required.
    \item[\textsc{RT-PROJ}] Then $r = \proj{r'}{i}$ and $\gt = \gt_i$.

      By inversion, $\tc{\gS;\gY}{\addr{\ga}{r'}}{\subrec{\gt}{i}{n}}$.

      By induction, $\Read{H}{\addr{\ga}{r'}}{l}$
      and $\tc{\gS;\gY}{l}{\subrec{\gt}{i}{n}}$ for some $l$.

      There is only one rule (\textsc{LT-REC}) for deriving $\tc{\gS;\gY}{l}{\subrec{\gt}{i}{n}}$.

      Thus, by inversion, $l = \subrec{l'}{j}{n}$ 
      and $\forall j.~\tc{\gS;\gY}{l'_j}{\gt_j}$.
      Specifically, $\tc{\gS;\gY}{l'_i}{\gt_i}$.

      Then by \textsc{RD-PROJ}, we have $\Read{H}{\addr{\ga}{\proj{r'}{i}}}{l'_i}$.
    \item[\textsc{RT-PAY}] Then $r = \pay{r'}$ and $\gt = \gt_i$.

      By inversion, $\gY(\ga,r')=i$ and $\tc{\gS;\gY}{\addr{\ga}{r'}}{\subvar{\gt}{i}{n}}$.

      By induction, $\Read{H}{\addr{\ga}{r'}}{l}$
      and $\tc{\gS;\gY}{l}{\subvar{\gt}{i}{n}}$.

      From $\tc{}{H}{\gS;\gY}$, we know that because $\gY(\ga,r')=i$
      it follows that $\Read{H}{\addr{\ga}{r'}}{\varval{i}{l'}}$ for some $l'$.

      By Read Uniqueness, we know that $l = \varval{i}{l'}$.

      There is only one rule (\textsc{LT-VAR}) for deriving 
      $\tc{\gS;\gY}{\varval{i}{l'}}{\subvar{\gt}{i}{n}}$.

      Thus, by inversion, $\tc{\gS;\gY}{l'}{\gt_i}$.

      Then by \textsc{RD-PAY}, we have $\Read{H}{\addr{\ga}{\pay{r'}}}{l'}$.
    \item[\textsc{RT-UNROLL}] 
      Then $r = \unroll{\tyfix{X}{\gt'}}{r'}$ and $\gt = \sub{X}{\tyfix{X}{\gt'}}{\gt'}$.

      By inversion, $\tc{\gS;\gY}{\addr{\ga}{r'}}{\tyfix{X}{\gt'}}$.

      By induction, $\Read{H}{\addr{\ga}{r'}}{l}$ 
      and $\tc{\gS;\gY}{l}{\tyfix{X}{\gt'}}$ for some $l$.

      There is only one rule (\textsc{LT-ROLL}) for deriving $\tc{\gS;\gY}{l}{\tyfix{X}{\gt'}}$.

      Thus, by inversion, $l = \roll{\tyfix{X}{\gt'}}{l'}$
      and $\tc{\gS;\gY}{l'}{\sub{X}{\tyfix{X}{\gt'}}{\gt'}}$.

      Then by \textsc{RD-UNROLL}, we have
      $\Read{H}{\addr{\ga}{\unroll{\tyfix{X}{\gt'}}{r'}}}{l'}$.
  \end{itemize}
\end{proof}

\section*{Path Type Preservation}
The final piece we need for preservation is how to relate the typing context $\gG$,
the variable map $V$, and the heap type $\gS$. This is simply the requirement that
the map $V$ preserves typing.

\begin{mathpar}
  \infer
  {\forall x \in \dom{\gG}.~\gG(x)=\gS(V(x))}
  {\gG;V\vdash\gS}
\end{mathpar}

Finally, we can state and prove path type preservation.
Under well-formed heap and contexts, typed paths evaluate to routes of the same type.

\begin{lem}[Path Type Preservation]
  If $\tc{}{H}{\gS;\gY}$, $\gG;V\vdash\gS$, $\tc{\gG}{\lv{x}{p}}{\gt}$,
  and $\ev{V;H}{\lv{x}{p}}{\addr{\ga}{r}}$, then $\tc{\gS;\gY}{\addr{\ga}{r}}{\gt}$.
\end{lem}

\begin{proof}[Path Type Preservation]
  We will use induction on the derivation of $\ev{V;H}{\lv{x}{p}}{\addr{\ga}{r}}$.
  \begin{itemize}
    \item[PE-BASE] Then $p = \base$ and $r = \base$.

      By inversion, $V(x)=\ga$.

      There is only one rule (\textsc{PT-BASE}) for deriving $\tc{\gG}{\lv{x}{\base}}{\gt}$.

      Thus, by inversion, we have $\gG(x)=\gt$.

      From $\gG;V\vdash\gS$, we know $\gG(x)=\gS(V(x))$. Thus, $\gt=\gS(\ga)$.

      Then by \textsc{RT-BASE}, we have $\tc{\gS;\gY}{\addr{\ga}{\base}}{\gt}$.
    \item[PE-DEOWN] Then $p = \deref{p'}$.

      By inversion, $\ev{V;H}{\lv{x}{p'}}{\addr{\ga'}{r'}}$
      and $\Read{H}{\addr{\ga'}{r'}}{\own{\addr{\ga}{r}}}$.

      There are two possible rules for deriving $\tc{\gG}{\lv{x}{\deref{p'}}}{\gt}$.
      \begin{itemize}
	\item[PT-DEOWN]

	  By induction, $\tc{\gS;\gY}{\addr{\ga'}{r'}}{\own{\gt}}$.

	  By Read Safety, we have $\Read{H}{\addr{\ga'}{r'}}{l}$
	  and $\tc{\gS;\gY}{l}{\own{\gt}}$ for some $l$.

	  By Read Uniqueness, we have $l =\ \own{\addr{\ga}{r}}$.

	  There is only one rule (\textsc{LT-OWN}) for deriving 
	  $\tc{\gS;\gY}{\own{\addr{\ga}{r}}}{\own{\gt}}$.

	  Thus, by inversion, $\tc{\gS;\gY}{\addr{\ga}{r}}{\gt}$.
	\item[PT-DEREF]

	  By inversion, $\tc{\gG}{\lv{x}{p'}}{\tyref{\lt}{q}{\gt}}$.

	  By induction, $\tc{\gS;\gY}{\addr{\ga'}{r'}}{\tyref{\lt}{q}{\gt}}$.

	  By Read Safety, we have $\Read{H}{\addr{\ga'}{r'}}{l}$
	  and $\tc{\gS;\gY}{l}{\tyref{\lt}{q}{\gt}}$ for some $l$.

	  By Read Uniqueness, we have $l =\ \own{\addr{\ga}{r}}$.

	  There are no rules for deriving
	  $\tc{\gS;\gY}{\own{\addr{\ga}{r}}}{\tyref{\lt}{q}{\gt}}$.

	  Thus, by inversion, this case is impossible.
      \end{itemize}
    \item[PE-DEREF] 
      Similar to the \textsc{PE-DEOWN} case, but switching the roles of
      \textsc{PT-DEOWN} and \textsc{PT-DEREF}.
    \item[PE-PROJ] Then $p = \proj{p'}{i}$ and $r = \proj{r'}{i}$.

      By inversion, $\ev{V;H}{\lv{x}{p'}}{\addr{\ga}{r'}}$.

      There is only one rule (\textsc{PT-PROJ}) for deriving $\tc{\gG}{\lv{x}{\proj{p'}{i}}}{\gt}$.

      Thus, by inversion, we have $\gt = \gt_i$ and $\tc{\gG}{\lv{x}{p'}}{\subrec{\gt}{i}{n}}$.

      By induction, we have $\tc{\gS;\gY}{\addr{\ga}{r'}}{\subrec{\gt}{i}{n}}$.

      Then by \textsc{RT-PROJ}, we have $\tc{\gS;\gY}{\addr{\ga}{\proj{r'}{i}}}{\gt_i}$.
    \item[PE-UNROLL] Then $p = \unroll{\gt'}{p'}$ and $r = \unroll{\gt'}{r'}$.

      By inversion, $\ev{V;H}{\lv{x}{p'}}{\addr{\ga}{r'}}$.

      There is only one rule (\textsc{PT-UNROLL}) for deriving
      $\tc{\gG}{\lv{x}{\unroll{\gt'}{p'}}}{\gt}$.

      Thus, by inversion, we have $\gt' = \tyfix{X}{\gt''}$,
      $\gt = \sub{X}{\tyfix{X}{\gt''}}{\gt''}$, and
      $\tc{\gG}{\lv{x}{p'}}{\tyfix{X}{\gt''}}$.

      By induction, we have $\tc{\gS;\gY}{\addr{\ga}{r'}}{\tyfix{X}{\gt''}}$.

      Then by \textsc{RT-UNROLL}, we have
      $\tc{\gS;\gY}{\addr{\ga}{\unroll{\tyfix{X}{\gt''}}{r'}}}{\sub{X}{\tyfix{X}{\gt''}}{\gt''}}$.
  \end{itemize}
\end{proof}
